---
title: "Simulations for Mixed Effects Models: Quantitative Microbiome Project Data"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: 
    toc: true
    theme: united
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      message = FALSE, comment = NA,
                      fig.width = 6.25, fig.height = 5)
library(openxlsx)
library(tidyverse)
library(mia)
library(ggpubr)
library(doRNG)
library(doParallel)
library(magrittr)
library(qwraps2)
library(ANCOMBC)
library(MicrobiomeStat)
```

```{r helper}
options(qwraps2_markup = "markdown")
summary_template =
  list("Power" =
         list("power_mean" = ~ round(mean(X1, na.rm = TRUE), 2),
              "power_sd" = ~ round(sd(X1, na.rm = TRUE), 2)),
       "FDR" = 
         list("fdr_mean" = ~ round(mean(X2, na.rm = TRUE), 2),
              "fdr_sd" = ~ round(sd(X2, na.rm = TRUE), 2))
  )

wb = createWorkbook()
```

# Case 1: random intercept models {.tabset}

Fixed effects: 

  + 1 categorical exposure with three levels: cat_cov

  + 1 continuous confounder: cont_cov 

Random effects: 

  + Random intercept: group with 3 levels and standard deviation 1

```{r}
# Simulation settings
set.seed(12345)
data(QMP, package = "ANCOMBC")
n = c(30, 60, 90, 150, 300)
d = ncol(QMP)
diff_prop = c(0.1, 0.2, 0.5, 0.9)
iter_num = 100
seed = seq_len(iter_num)
df_sim_params = data.frame(expand.grid(n, diff_prop, seed)) %>%
  dplyr::rename(n = Var1, diff_prop = Var2, seed = Var3) %>%
  arrange(n, diff_prop, seed)
list_sim_params = apply(df_sim_params, 1, paste0, collapse = "_")

lfc_value = c(-2, -1, 1, 2)
lfc_cat2_vs_1_list = vector("list", length = length(diff_prop))
for (i in seq_along(diff_prop)) {
  lfc_cat2_vs_1_list[[i]] = sample(c(0, lfc_value), size = d, replace = TRUE,
                                   prob = c(1 - diff_prop[i],
                                            rep(diff_prop[i]/length(lfc_value), length(lfc_value))))
}
names(lfc_cat2_vs_1_list) = diff_prop

lfc_cat3_vs_1_list = vector("list", length = length(diff_prop))
for (i in seq_along(diff_prop)) {
  lfc_cat3_vs_1_list[[i]] = sample(c(0, lfc_value), size = d, replace = TRUE,
                                   prob = c(1 - diff_prop[i],
                                            rep(diff_prop[i]/length(lfc_value), length(lfc_value))))
}
names(lfc_cat3_vs_1_list) = diff_prop

lfc_cont_list = vector("list", length = length(diff_prop))
for (i in seq_along(diff_prop)) {
  lfc_cont_list[[i]] = sample(c(0, 1), size = d, replace = TRUE,
                              prob = c(1 - diff_prop[i], diff_prop[i]))
}
names(lfc_cont_list) = diff_prop
```

## ANCOM-BC2

```{r, eval=FALSE}
# detect all cpus per task
ncpus = Sys.getenv("SLURM_CPUS_PER_TASK")
# make cluster with ncpus
cl = makeCluster(strtoi(ncpus))
registerDoParallel(cl)

res_sim = foreach(i = list_sim_params, .combine = rbind, .verbose = TRUE, 
                  .packages = c("ANCOMBC", "tidyverse")) %dorng% 
  {
    params = strsplit(i, "_")[[1]]
    n = as.numeric(params[1])
    diff_prop = as.numeric(params[2])
    seed = as.numeric(params[3])
    
    # Generate the true abundances
    set.seed(seed)
    abn_data = sim_plnm(abn_table = QMP, taxa_are_rows = FALSE, prv_cut = 0.05, 
                        n = n, lib_mean = 1e8, disp = 0.5)
    log_abn_data = log(abn_data + 1e-5)
    
    # Generate the sample and feature meta data
    # Sampling fractions are set to differ by batches
    smd = data.frame(sample = paste0("S", seq_len(n)),
                     samp_frac = log(c(runif(n/3, min = 1e-4, max = 1e-3),
                                       runif(n/3, min = 1e-3, max = 1e-2),
                                       runif(n/3, min = 1e-2, max = 1e-1))),
                     cont_cov = rnorm(n),
                     cat_cov = as.factor(rep(seq_len(3), each = n/3)),
                     batch = rep(c("A", "B", "C"), n/3))
    
    d = nrow(abn_data)   
    lfc_cat2_vs_1 = lfc_cat2_vs_1_list[[as.character(diff_prop)]]
    lfc_cat3_vs_1 = lfc_cat3_vs_1_list[[as.character(diff_prop)]]
    lfc_cont = lfc_cont_list[[as.character(diff_prop)]]
    fmd = data.frame(taxon = paste0("T", seq_len(d)),
                     seq_eff = log(runif(d, min = 0.1, max = 1)),
                     lfc_cont = lfc_cont,
                     lfc_cat2_vs_1 = lfc_cat2_vs_1,
                     lfc_cat3_vs_1 = lfc_cat3_vs_1)
    
    # Add fixed effects
    smd_dmy = model.matrix(~ 0 + cont_cov + cat_cov, data = smd)
    log_abn_data = log_abn_data + outer(fmd$lfc_cont, smd_dmy[, "cont_cov"] )
    log_abn_data = log_abn_data + outer(fmd$lfc_cat2_vs_1, smd_dmy[, "cat_cov2"])
    log_abn_data = log_abn_data + outer(fmd$lfc_cat3_vs_1, smd_dmy[, "cat_cov3"])
    
    # Add random effects
    rand_eff = rnorm(3, mean = 0, sd = 1)
    rand_int_mat = model.matrix(~ 0 + batch, data = smd)
    log_abn_data = t(apply(log_abn_data , 1, function(x) x + rand_int_mat %*% rand_eff))
    rownames(log_abn_data) = paste0("T", seq_len(d))
    colnames(log_abn_data) = paste0("S", seq_len(n))
    
    # Add sample- and taxon-specific biases
    log_otu_data = t(t(log_abn_data) + smd$samp_frac)
    log_otu_data = log_otu_data + fmd$seq_eff
    otu_data = round(exp(log_otu_data))
    
    # Create the tse object
    assays = S4Vectors::SimpleList(counts = otu_data)
    smd = S4Vectors::DataFrame(smd)
    tse = TreeSummarizedExperiment::TreeSummarizedExperiment(assays = assays, colData = smd)
    
    # Run ANCOM-BC2
    set.seed(123)
    output = ancombc2(data = tse, assay_name = "counts", tax_level = NULL,
                      fix_formula = "cont_cov + cat_cov", 
                      rand_formula = "(1 | batch)",
                      p_adj_method = "holm", 
                      prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                      group = "cat_cov", struc_zero = FALSE, neg_lb = FALSE,
                      alpha = 0.05, n_cl = 1, verbose = FALSE,
                      global = FALSE, pairwise = FALSE, 
                      dunnet = TRUE, trend = FALSE,
                      iter_control = list(tol = 1e-2, max_iter = 10, 
                                          verbose = FALSE),
                      em_control = list(tol = 1e-5, max_iter = 100),
                      lme_control = lme4::lmerControl(), 
                      mdfdr_control = list(fwer_ctrl_method = "holm", B = 100))
    
    res_dunn = output$res_dunn
    
    res_merge1 = res_dunn %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = lfc_cat_cov2 * diff_cat_cov2,
                       lfc_est2 = lfc_cat_cov3 * diff_cat_cov3) %>%
      dplyr::left_join(fmd %>%
                         dplyr::transmute(taxon, 
                                          lfc_true1 = lfc_cat2_vs_1,
                                          lfc_true2 = lfc_cat3_vs_1),
                       by = "taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = case_when(lfc_est1 > 0 ~ 1,
                                            lfc_est1 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_est2 = case_when(lfc_est2 > 0 ~ 1,
                                            lfc_est2 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_true1 = case_when(lfc_true1 > 0 ~ 1,
                                             lfc_true1 < 0 ~ -1,
                                             TRUE ~ 0),
                       lfc_true2 = case_when(lfc_true2 > 0 ~ 1,
                                             lfc_true2 < 0 ~ -1,
                                             TRUE ~ 0))
    lfc_est1 = res_merge1$lfc_est1
    lfc_true1 = res_merge1$lfc_true1
    lfc_est2 = res_merge1$lfc_est2
    lfc_true2 = res_merge1$lfc_true2
    tp1 = sum(lfc_true1 == 1 & lfc_est1 == 1) +
      sum(lfc_true1 == -1 & lfc_est1 == -1)
    fp1 = sum(lfc_true1 == 0 & lfc_est1 != 0) +
      sum(lfc_true1 == 1 & lfc_est1 == -1) +
      sum(lfc_true1 == -1 & lfc_est1 == 1)
    fn1 = sum(lfc_true1 != 0 & lfc_est1 == 0)
    
    tp2 = sum(lfc_true2 == 1 & lfc_est2 == 1) +
      sum(lfc_true2 == -1 & lfc_est2 == -1)
    fp2 = sum(lfc_true2 == 0 & lfc_est2 != 0) +
      sum(lfc_true2 == 1 & lfc_est2 == -1) +
      sum(lfc_true2 == -1 & lfc_est2 == 1)
    fn2 = sum(lfc_true2 != 0 & lfc_est2 == 0)
    tp = tp1 + tp2
    fp = fp1 + fp2
    fn = fn1 + fn2
    power1 = tp/(tp + fn)
    fdr1 = fp/(tp + fp)
    
    res_merge2 = res_dunn %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = lfc_cat_cov2 * diff_cat_cov2 * passed_ss_cat_cov2,
                       lfc_est2 = lfc_cat_cov3 * diff_cat_cov3 * passed_ss_cat_cov3) %>%
      dplyr::left_join(fmd %>%
                         dplyr::transmute(taxon, 
                                          lfc_true1 = lfc_cat2_vs_1,
                                          lfc_true2 = lfc_cat3_vs_1),
                       by = "taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = case_when(lfc_est1 > 0 ~ 1,
                                            lfc_est1 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_est2 = case_when(lfc_est2 > 0 ~ 1,
                                            lfc_est2 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_true1 = case_when(lfc_true1 > 0 ~ 1,
                                             lfc_true1 < 0 ~ -1,
                                             TRUE ~ 0),
                       lfc_true2 = case_when(lfc_true2 > 0 ~ 1,
                                             lfc_true2 < 0 ~ -1,
                                             TRUE ~ 0))
    lfc_est1 = res_merge2$lfc_est1
    lfc_true1 = res_merge2$lfc_true1
    lfc_est2 = res_merge2$lfc_est2
    lfc_true2 = res_merge2$lfc_true2
    tp1 = sum(lfc_true1 == 1 & lfc_est1 == 1) +
      sum(lfc_true1 == -1 & lfc_est1 == -1)
    fp1 = sum(lfc_true1 == 0 & lfc_est1 != 0) +
      sum(lfc_true1 == 1 & lfc_est1 == -1) +
      sum(lfc_true1 == -1 & lfc_est1 == 1)
    fn1 = sum(lfc_true1 != 0 & lfc_est1 == 0)
    
    tp2 = sum(lfc_true2 == 1 & lfc_est2 == 1) +
      sum(lfc_true2 == -1 & lfc_est2 == -1)
    fp2 = sum(lfc_true2 == 0 & lfc_est2 != 0) +
      sum(lfc_true2 == 1 & lfc_est2 == -1) +
      sum(lfc_true2 == -1 & lfc_est2 == 1)
    fn2 = sum(lfc_true2 != 0 & lfc_est2 == 0)
    tp = tp1 + tp2
    fp = fp1 + fp2
    fn = fn1 + fn2
    power2 = tp/(tp + fn)
    fdr2 = fp/(tp + fp)
    
    c(power1, fdr1, power2, fdr2)
  }

stopCluster(cl)

write_csv(data.frame(res_sim), "qmp_sim_int_ancombc2.csv")
```

## LinDA

```{r, eval=FALSE}
cl = makeCluster(4)
registerDoParallel(cl)

res_sim = foreach(i = list_sim_params, .combine = rbind, .verbose = TRUE, 
                  .packages = c("ANCOMBC", "tidyverse", "MicrobiomeStat")) %dorng% 
  {
    params = strsplit(i, "_")[[1]]
    n = as.numeric(params[1])
    diff_prop = as.numeric(params[2])
    seed = as.numeric(params[3])
    
    # Generate the true abundances
    set.seed(seed)
    abn_data = sim_plnm(abn_table = QMP, taxa_are_rows = FALSE, prv_cut = 0.05, 
                        n = n, lib_mean = 1e8, disp = 0.5)
    log_abn_data = log(abn_data + 1e-5)
    
    # Generate the sample and feature meta data
    # Sampling fractions are set to differ by batches
    smd = data.frame(sample = paste0("S", seq_len(n)),
                     samp_frac = log(c(runif(n/3, min = 1e-4, max = 1e-3),
                                       runif(n/3, min = 1e-3, max = 1e-2),
                                       runif(n/3, min = 1e-2, max = 1e-1))),
                     cont_cov = rnorm(n),
                     cat_cov = as.factor(rep(seq_len(3), each = n/3)),
                     batch = rep(c("A", "B", "C"), n/3))
    
    d = nrow(abn_data)   
    lfc_cat2_vs_1 = lfc_cat2_vs_1_list[[as.character(diff_prop)]]
    lfc_cat3_vs_1 = lfc_cat3_vs_1_list[[as.character(diff_prop)]]
    lfc_cont = lfc_cont_list[[as.character(diff_prop)]]
    fmd = data.frame(taxon = paste0("T", seq_len(d)),
                     seq_eff = log(runif(d, min = 0.1, max = 1)),
                     lfc_cont = lfc_cont,
                     lfc_cat2_vs_1 = lfc_cat2_vs_1,
                     lfc_cat3_vs_1 = lfc_cat3_vs_1)
    
    # Add fixed effects
    smd_dmy = model.matrix(~ 0 + cont_cov + cat_cov, data = smd)
    log_abn_data = log_abn_data + outer(fmd$lfc_cont, smd_dmy[, "cont_cov"] )
    log_abn_data = log_abn_data + outer(fmd$lfc_cat2_vs_1, smd_dmy[, "cat_cov2"])
    log_abn_data = log_abn_data + outer(fmd$lfc_cat3_vs_1, smd_dmy[, "cat_cov3"])
    
    # Add random effects
    rand_eff = rnorm(3, mean = 0, sd = 1)
    rand_int_mat = model.matrix(~ 0 + batch, data = smd)
    log_abn_data = t(apply(log_abn_data , 1, function(x) x + rand_int_mat %*% rand_eff))
    rownames(log_abn_data) = paste0("T", seq_len(d))
    colnames(log_abn_data) = paste0("S", seq_len(n))
    
    # Add sample- and taxon-specific biases
    log_otu_data = t(t(log_abn_data) + smd$samp_frac)
    log_otu_data = log_otu_data + fmd$seq_eff
    otu_data = round(exp(log_otu_data))
    
    # Remove samples with low library sizes
    idx = which(colSums(otu_data) > 1000)
    otu_data = otu_data[, idx]
    smd = smd[idx, ]
    
    # Run LinDA
    output = linda(feature.dat = otu_data, meta.dat = smd,
                   formula = "~ cont_cov + cat_cov + (1 | batch)",
                   alpha = 0.05, 
                   prev.filter = 0.10, 
                   mean.abund.filter = 0,
                   adaptive = TRUE,
                   max.abund.filter = 0,
                   p.adj.method = "holm",
                   n.cores = 1, 
                   verbose = FALSE)
    
    res = output$output
    res_merge = res$cat_cov2 %>%
      rownames_to_column("taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = log2FoldChange * reject) %>%
      dplyr::left_join(fmd %>%
                         dplyr::transmute(taxon, 
                                          lfc_true1 = lfc_cat2_vs_1),
                       by = "taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = case_when(lfc_est1 > 0 ~ 1,
                                            lfc_est1 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_true1 = case_when(lfc_true1 > 0 ~ 1,
                                             lfc_true1 < 0 ~ -1,
                                             TRUE ~ 0)) %>%
      dplyr::left_join(
        res$cat_cov3 %>%
          rownames_to_column("taxon") %>%
          dplyr::transmute(taxon, 
                           lfc_est2 = log2FoldChange * reject) %>%
          dplyr::left_join(fmd %>%
                             dplyr::transmute(taxon, 
                                              lfc_true2 = lfc_cat3_vs_1),
                           by = "taxon") %>%
          dplyr::transmute(taxon, 
                           lfc_est2 = case_when(lfc_est2 > 0 ~ 1,
                                                lfc_est2 < 0 ~ -1,
                                                TRUE ~ 0),
                           lfc_true2 = case_when(lfc_true2 > 0 ~ 1,
                                                 lfc_true2 < 0 ~ -1,
                                                 TRUE ~ 0)),
        by = "taxon")
    
    lfc_est1 = res_merge$lfc_est1
    lfc_true1 = res_merge$lfc_true1
    lfc_est2 = res_merge$lfc_est2
    lfc_true2 = res_merge$lfc_true2
    
    tp1 = sum(lfc_true1 != 0 & lfc_est1 != 0)
    fp1 = sum(lfc_true1 == 0 & lfc_est1 != 0)
    fn1 = sum(lfc_true1 != 0 & lfc_est1 == 0)
    tp2 = sum(lfc_true2 != 0 & lfc_est2 != 0)
    fp2 = sum(lfc_true2 == 0 & lfc_est2 != 0)
    fn2 = sum(lfc_true2 != 0 & lfc_est2 == 0)
    tp = tp1 + tp2
    fp = fp1 + fp2
    fn = fn1 + fn2
    power = tp/(tp + fn)
    fdr = fp/(tp + fp)
    
    c(power, fdr)
  }

stopCluster(cl)

write_csv(data.frame(res_sim), "qmp_sim_int_linda.csv")
```

## Visualization

```{r, fig.width=10, fig.height=6.25}
df_ancombc2 = read_csv("../data/qmp/sim_mixed/qmp_sim_int_ancombc2.csv")
df_linda = read_csv("../data/qmp/sim_mixed/qmp_sim_int_linda.csv")

simpattern = distinct(df_sim_params, n, diff_prop) %>%
  unite("setting", n:diff_prop, sep = ", ")

df_ancombc2_no_filter = df_ancombc2 %>%
  dplyr::select(X1, X2) %>%
  mutate(method = "ANCOM-BC2 (No Filter)",
         setting = rep(simpattern$setting, each = iter_num))
df_ancombc2_ss_filter = df_ancombc2 %>%
  dplyr::transmute(X1 = X3, X2 = X4) %>%
  mutate(method = "ANCOM-BC2 (SS Filter)",
         setting = rep(simpattern$setting, each = iter_num))
df_linda = df_linda %>%
  mutate(method = "LinDA",
         setting = rep(simpattern$setting, each = iter_num))

df_fig = df_ancombc2_no_filter %>%
  bind_rows(df_ancombc2_ss_filter) %>%
  bind_rows(df_linda) %>%
  separate(setting, c("n", "diff_prop"), ", ")
df_fig$method = factor(df_fig$method, 
                       levels = c("ANCOM-BC2 (No Filter)", 
                                  "ANCOM-BC2 (SS Filter)", 
                                  "LinDA"))

fig_power_int = df_fig %>%
  ggline(x = "n", y = "X1", add = "mean_se",
         color = "method", xlab = "Sample Size", ylab = "Power", 
         facet.by = "diff_prop", nrow = 1) +
  scale_color_manual(values = c("#E64B35FF", "#4DBBD5FF", "#3C5488FF")) +
  scale_x_discrete(labels = c(10, 20, 30, 50, 100)) +
  guides(color = guide_legend(title = NULL))

fig_fdr_int = df_fig %>%
  ggline(x = "n", y = "X2", add = "mean_se",
         color = "method", xlab = "Sample Size", ylab = "mdFDR", 
         facet.by = "diff_prop", nrow = 1) +
  scale_color_manual(values = c("#E64B35FF", "#4DBBD5FF", "#3C5488FF")) +
  scale_x_discrete(labels = c(10, 20, 30, 50, 100)) +
  guides(color = guide_legend(title = NULL)) +
  geom_hline(yintercept = 0.05, linetype = "dashed")

fig_int = ggarrange(fig_fdr_int, fig_power_int, 
                    ncol = 1, common.legend = TRUE)

# Simulation summary
df_tab = df_fig %>%
  unite("setting", n:diff_prop, sep = ", ")
df_tab$setting = factor(df_tab$setting, 
                        levels = c(paste0("30, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("60, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("90, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("150, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("300, ", c(0.1, 0.2, 0.5, 0.9))))

tab = df_tab %>% 
  group_by(method, setting) %>% 
  summary_table(summary_template)

addWorksheet(wb, "int")
writeData(wb, "int", tab)
```

# Case 2: random coefficients models {.tabset}

Fixed effects: 
  
  + 1 categorical exposure with three levels: cat_cov
  
  + 1 continuous confounder: cont_cov 

Random effects: 

  + Random intercept: group with 3 levels and standard deviation 1
  
  + Random slope: cont_cov with standard deviation 1.5
  
  + A correlation of 0.5 between random intercept and random slope

```{r}
# Simulation settings
set.seed(12345)
data(QMP, package = "ANCOMBC")
n = c(30, 60, 90, 150, 300)
d = ncol(QMP)
diff_prop = c(0.1, 0.2, 0.5, 0.9)
iter_num = 100
seed = seq_len(iter_num)
df_sim_params = data.frame(expand.grid(n, diff_prop, seed)) %>%
  dplyr::rename(n = Var1, diff_prop = Var2, seed = Var3) %>%
  arrange(n, diff_prop, seed)
list_sim_params = apply(df_sim_params, 1, paste0, collapse = "_")

lfc_value = c(-2, -1, 1, 2)
lfc_cat2_vs_1_list = vector("list", length = length(diff_prop))
for (i in seq_along(diff_prop)) {
  lfc_cat2_vs_1_list[[i]] = sample(c(0, lfc_value), size = d, replace = TRUE,
                                   prob = c(1 - diff_prop[i],
                                            rep(diff_prop[i]/length(lfc_value), length(lfc_value))))
}
names(lfc_cat2_vs_1_list) = diff_prop

lfc_cat3_vs_1_list = vector("list", length = length(diff_prop))
for (i in seq_along(diff_prop)) {
  lfc_cat3_vs_1_list[[i]] = sample(c(0, lfc_value), size = d, replace = TRUE,
                                   prob = c(1 - diff_prop[i],
                                            rep(diff_prop[i]/length(lfc_value), length(lfc_value))))
}
names(lfc_cat3_vs_1_list) = diff_prop

lfc_cont_list = vector("list", length = length(diff_prop))
for (i in seq_along(diff_prop)) {
  lfc_cont_list[[i]] = sample(c(0, 1), size = d, replace = TRUE,
                              prob = c(1 - diff_prop[i], diff_prop[i]))
}
names(lfc_cont_list) = diff_prop
```

## ANCOM-BC2

```{r, eval=FALSE}
# detect all cpus per task
ncpus = Sys.getenv("SLURM_CPUS_PER_TASK")
# make cluster with ncpus
cl = makeCluster(strtoi(ncpus))
registerDoParallel(cl)

res_sim = foreach(i = list_sim_params, .combine = rbind, .verbose = TRUE, 
                  .packages = c("ANCOMBC", "tidyverse")) %dorng% 
  {
    params = strsplit(i, "_")[[1]]
    n = as.numeric(params[1])
    diff_prop = as.numeric(params[2])
    seed = as.numeric(params[3])
    
    # Generate the true abundances
    set.seed(seed)
    abn_data = sim_plnm(abn_table = QMP, taxa_are_rows = FALSE, prv_cut = 0.05, 
                        n = n, lib_mean = 1e8, disp = 0.5)
    log_abn_data = log(abn_data + 1e-5)
    
    # Generate the sample and feature meta data
    # Sampling fractions are set to differ by batches
    smd = data.frame(sample = paste0("S", seq_len(n)),
                     samp_frac = log(c(runif(n/3, min = 1e-4, max = 1e-3),
                                       runif(n/3, min = 1e-3, max = 1e-2),
                                       runif(n/3, min = 1e-2, max = 1e-1))),
                     cont_cov = rnorm(n),
                     cat_cov = as.factor(rep(seq_len(3), each = n/3)),
                     batch = rep(c("A", "B", "C"), n/3))
    
    d = nrow(abn_data)   
    lfc_cat2_vs_1 = lfc_cat2_vs_1_list[[as.character(diff_prop)]]
    lfc_cat3_vs_1 = lfc_cat3_vs_1_list[[as.character(diff_prop)]]
    lfc_cont = lfc_cont_list[[as.character(diff_prop)]]
    fmd = data.frame(taxon = paste0("T", seq_len(d)),
                     seq_eff = log(runif(d, min = 0.1, max = 1)),
                     lfc_cont = lfc_cont,
                     lfc_cat2_vs_1 = lfc_cat2_vs_1,
                     lfc_cat3_vs_1 = lfc_cat3_vs_1)
    
    # Add fixed effects
    smd_dmy = model.matrix(~ 0 + cont_cov + cat_cov, data = smd)
    log_abn_data = log_abn_data + outer(fmd$lfc_cont, smd_dmy[, "cont_cov"] )
    log_abn_data = log_abn_data + outer(fmd$lfc_cat2_vs_1, smd_dmy[, "cat_cov2"])
    log_abn_data = log_abn_data + outer(fmd$lfc_cat3_vs_1, smd_dmy[, "cat_cov3"])
    
    # Add random effects
    covmat =  matrix(c(1, 0.5, 0.5, 1.5), 2, 2)
    rand_eff = MASS::mvrnorm(n = 3, mu = c(0, 0), Sigma = covmat)
    rand_int_mat = model.matrix(~ 0 + batch, data = smd)
    rand_slope_mat = model.matrix(~ 0 + batch, data = smd) * smd$cont_cov
    log_abn_data = t(apply(log_abn_data , 1, 
                           function(x) 
                             x + rand_int_mat %*% rand_eff[, 1] + rand_slope_mat %*% rand_eff[, 2]))
    rownames(log_abn_data) = paste0("T", seq_len(d))
    colnames(log_abn_data) = paste0("S", seq_len(n))
    
    # Add sample- and taxon-specific biases
    log_otu_data = t(t(log_abn_data) + smd$samp_frac)
    log_otu_data = log_otu_data + fmd$seq_eff
    otu_data = round(exp(log_otu_data))
    
    # Create the tse object
    assays = S4Vectors::SimpleList(counts = otu_data)
    smd = S4Vectors::DataFrame(smd)
    tse = TreeSummarizedExperiment::TreeSummarizedExperiment(assays = assays, colData = smd)
    
    # Run ANCOM-BC2
    set.seed(123)
    output = ancombc2(data = tse, assay_name = "counts", tax_level = NULL,
                      fix_formula = "cont_cov + cat_cov", 
                      rand_formula = "(cont_cov | batch)",
                      p_adj_method = "holm", 
                      prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                      group = "cat_cov", struc_zero = FALSE, neg_lb = FALSE,
                      alpha = 0.05, n_cl = 1, verbose = FALSE,
                      global = FALSE, pairwise = FALSE, 
                      dunnet = TRUE, trend = FALSE,
                      iter_control = list(tol = 1e-2, max_iter = 10, 
                                          verbose = FALSE),
                      em_control = list(tol = 1e-5, max_iter = 100),
                      lme_control = lme4::lmerControl(), 
                      mdfdr_control = list(fwer_ctrl_method = "holm", B = 100))
    
    res_dunn = output$res_dunn
    
    res_merge1 = res_dunn %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = lfc_cat_cov2 * diff_cat_cov2,
                       lfc_est2 = lfc_cat_cov3 * diff_cat_cov3) %>%
      dplyr::left_join(fmd %>%
                         dplyr::transmute(taxon, 
                                          lfc_true1 = lfc_cat2_vs_1,
                                          lfc_true2 = lfc_cat3_vs_1),
                       by = "taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = case_when(lfc_est1 > 0 ~ 1,
                                            lfc_est1 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_est2 = case_when(lfc_est2 > 0 ~ 1,
                                            lfc_est2 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_true1 = case_when(lfc_true1 > 0 ~ 1,
                                             lfc_true1 < 0 ~ -1,
                                             TRUE ~ 0),
                       lfc_true2 = case_when(lfc_true2 > 0 ~ 1,
                                             lfc_true2 < 0 ~ -1,
                                             TRUE ~ 0))
    lfc_est1 = res_merge1$lfc_est1
    lfc_true1 = res_merge1$lfc_true1
    lfc_est2 = res_merge1$lfc_est2
    lfc_true2 = res_merge1$lfc_true2
    tp1 = sum(lfc_true1 == 1 & lfc_est1 == 1) +
      sum(lfc_true1 == -1 & lfc_est1 == -1)
    fp1 = sum(lfc_true1 == 0 & lfc_est1 != 0) +
      sum(lfc_true1 == 1 & lfc_est1 == -1) +
      sum(lfc_true1 == -1 & lfc_est1 == 1)
    fn1 = sum(lfc_true1 != 0 & lfc_est1 == 0)
    
    tp2 = sum(lfc_true2 == 1 & lfc_est2 == 1) +
      sum(lfc_true2 == -1 & lfc_est2 == -1)
    fp2 = sum(lfc_true2 == 0 & lfc_est2 != 0) +
      sum(lfc_true2 == 1 & lfc_est2 == -1) +
      sum(lfc_true2 == -1 & lfc_est2 == 1)
    fn2 = sum(lfc_true2 != 0 & lfc_est2 == 0)
    tp = tp1 + tp2
    fp = fp1 + fp2
    fn = fn1 + fn2
    power1 = tp/(tp + fn)
    fdr1 = fp/(tp + fp)
    
    res_merge2 = res_dunn %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = lfc_cat_cov2 * diff_cat_cov2 * passed_ss_cat_cov2,
                       lfc_est2 = lfc_cat_cov3 * diff_cat_cov3 * passed_ss_cat_cov3) %>%
      dplyr::left_join(fmd %>%
                         dplyr::transmute(taxon, 
                                          lfc_true1 = lfc_cat2_vs_1,
                                          lfc_true2 = lfc_cat3_vs_1),
                       by = "taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = case_when(lfc_est1 > 0 ~ 1,
                                            lfc_est1 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_est2 = case_when(lfc_est2 > 0 ~ 1,
                                            lfc_est2 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_true1 = case_when(lfc_true1 > 0 ~ 1,
                                             lfc_true1 < 0 ~ -1,
                                             TRUE ~ 0),
                       lfc_true2 = case_when(lfc_true2 > 0 ~ 1,
                                             lfc_true2 < 0 ~ -1,
                                             TRUE ~ 0))
    lfc_est1 = res_merge2$lfc_est1
    lfc_true1 = res_merge2$lfc_true1
    lfc_est2 = res_merge2$lfc_est2
    lfc_true2 = res_merge2$lfc_true2
    tp1 = sum(lfc_true1 == 1 & lfc_est1 == 1) +
      sum(lfc_true1 == -1 & lfc_est1 == -1)
    fp1 = sum(lfc_true1 == 0 & lfc_est1 != 0) +
      sum(lfc_true1 == 1 & lfc_est1 == -1) +
      sum(lfc_true1 == -1 & lfc_est1 == 1)
    fn1 = sum(lfc_true1 != 0 & lfc_est1 == 0)
    
    tp2 = sum(lfc_true2 == 1 & lfc_est2 == 1) +
      sum(lfc_true2 == -1 & lfc_est2 == -1)
    fp2 = sum(lfc_true2 == 0 & lfc_est2 != 0) +
      sum(lfc_true2 == 1 & lfc_est2 == -1) +
      sum(lfc_true2 == -1 & lfc_est2 == 1)
    fn2 = sum(lfc_true2 != 0 & lfc_est2 == 0)
    tp = tp1 + tp2
    fp = fp1 + fp2
    fn = fn1 + fn2
    power2 = tp/(tp + fn)
    fdr2 = fp/(tp + fp)
    
    c(power1, fdr1, power2, fdr2)
  }

stopCluster(cl)

write_csv(data.frame(res_sim), "qmp_sim_slope_ancombc2.csv")
```

## LinDA

```{r, eval=FALSE}
cl = makeCluster(4)
registerDoParallel(cl)

res_sim = foreach(i = list_sim_params, .combine = rbind, .verbose = TRUE, 
                  .packages = c("ANCOMBC", "tidyverse", "MicrobiomeStat")) %dorng% 
  {
    params = strsplit(i, "_")[[1]]
    n = as.numeric(params[1])
    diff_prop = as.numeric(params[2])
    seed = as.numeric(params[3])
    
    # Generate the true abundances
    set.seed(seed)
    abn_data = sim_plnm(abn_table = QMP, taxa_are_rows = FALSE, prv_cut = 0.05, 
                        n = n, lib_mean = 1e8, disp = 0.5)
    log_abn_data = log(abn_data + 1e-5)
    
    # Generate the sample and feature meta data
    # Sampling fractions are set to differ by batches
    smd = data.frame(sample = paste0("S", seq_len(n)),
                     samp_frac = log(c(runif(n/3, min = 1e-4, max = 1e-3),
                                       runif(n/3, min = 1e-3, max = 1e-2),
                                       runif(n/3, min = 1e-2, max = 1e-1))),
                     cont_cov = rnorm(n),
                     cat_cov = as.factor(rep(seq_len(3), each = n/3)),
                     batch = rep(c("A", "B", "C"), n/3))
    
    d = nrow(abn_data)   
    lfc_cat2_vs_1 = lfc_cat2_vs_1_list[[as.character(diff_prop)]]
    lfc_cat3_vs_1 = lfc_cat3_vs_1_list[[as.character(diff_prop)]]
    lfc_cont = lfc_cont_list[[as.character(diff_prop)]]
    fmd = data.frame(taxon = paste0("T", seq_len(d)),
                     seq_eff = log(runif(d, min = 0.1, max = 1)),
                     lfc_cont = lfc_cont,
                     lfc_cat2_vs_1 = lfc_cat2_vs_1,
                     lfc_cat3_vs_1 = lfc_cat3_vs_1)
    
    # Add fixed effects
    smd_dmy = model.matrix(~ 0 + cont_cov + cat_cov, data = smd)
    log_abn_data = log_abn_data + outer(fmd$lfc_cont, smd_dmy[, "cont_cov"] )
    log_abn_data = log_abn_data + outer(fmd$lfc_cat2_vs_1, smd_dmy[, "cat_cov2"])
    log_abn_data = log_abn_data + outer(fmd$lfc_cat3_vs_1, smd_dmy[, "cat_cov3"])
    
    # Add random effects
    covmat =  matrix(c(1, 0.5, 0.5, 1.5), 2, 2)
    rand_eff = MASS::mvrnorm(n = 3, mu = c(0, 0), Sigma = covmat)
    rand_int_mat = model.matrix(~ 0 + batch, data = smd)
    rand_slope_mat = model.matrix(~ 0 + batch, data = smd) * smd$cont_cov
    log_abn_data = t(apply(log_abn_data , 1, 
                           function(x) 
                             x + rand_int_mat %*% rand_eff[, 1] + rand_slope_mat %*% rand_eff[, 2]))
    rownames(log_abn_data) = paste0("T", seq_len(d))
    colnames(log_abn_data) = paste0("S", seq_len(n))
    
    # Add sample- and taxon-specific biases
    log_otu_data = t(t(log_abn_data) + smd$samp_frac)
    log_otu_data = log_otu_data + fmd$seq_eff
    otu_data = round(exp(log_otu_data))
    
    # Remove samples with low library sizes
    idx = which(colSums(otu_data) > 1000)
    otu_data = otu_data[, idx]
    smd = smd[idx, ]
    
    # Run LinDA
    output = linda(feature.dat = otu_data, meta.dat = smd,
                   formula = "~ cont_cov + cat_cov + (cont_cov | batch)",
                   alpha = 0.05, 
                   prev.filter = 0.10, 
                   mean.abund.filter = 0,
                   adaptive = TRUE,
                   max.abund.filter = 0,
                   p.adj.method = "holm",
                   n.cores = 1, 
                   verbose = FALSE)
    
    res = output$output
    res_merge = res$cat_cov2 %>%
      rownames_to_column("taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = log2FoldChange * reject) %>%
      dplyr::left_join(fmd %>%
                         dplyr::transmute(taxon, 
                                          lfc_true1 = lfc_cat2_vs_1),
                       by = "taxon") %>%
      dplyr::transmute(taxon, 
                       lfc_est1 = case_when(lfc_est1 > 0 ~ 1,
                                            lfc_est1 < 0 ~ -1,
                                            TRUE ~ 0),
                       lfc_true1 = case_when(lfc_true1 > 0 ~ 1,
                                             lfc_true1 < 0 ~ -1,
                                             TRUE ~ 0)) %>%
      dplyr::left_join(
        res$cat_cov3 %>%
          rownames_to_column("taxon") %>%
          dplyr::transmute(taxon, 
                           lfc_est2 = log2FoldChange * reject) %>%
          dplyr::left_join(fmd %>%
                             dplyr::transmute(taxon, 
                                              lfc_true2 = lfc_cat3_vs_1),
                           by = "taxon") %>%
          dplyr::transmute(taxon, 
                           lfc_est2 = case_when(lfc_est2 > 0 ~ 1,
                                                lfc_est2 < 0 ~ -1,
                                                TRUE ~ 0),
                           lfc_true2 = case_when(lfc_true2 > 0 ~ 1,
                                                 lfc_true2 < 0 ~ -1,
                                                 TRUE ~ 0)),
        by = "taxon")
    
    lfc_est1 = res_merge$lfc_est1
    lfc_true1 = res_merge$lfc_true1
    lfc_est2 = res_merge$lfc_est2
    lfc_true2 = res_merge$lfc_true2
    
    tp1 = sum(lfc_true1 != 0 & lfc_est1 != 0)
    fp1 = sum(lfc_true1 == 0 & lfc_est1 != 0)
    fn1 = sum(lfc_true1 != 0 & lfc_est1 == 0)
    tp2 = sum(lfc_true2 != 0 & lfc_est2 != 0)
    fp2 = sum(lfc_true2 == 0 & lfc_est2 != 0)
    fn2 = sum(lfc_true2 != 0 & lfc_est2 == 0)
    tp = tp1 + tp2
    fp = fp1 + fp2
    fn = fn1 + fn2
    power = tp/(tp + fn)
    fdr = fp/(tp + fp)
    
    c(power, fdr)
  }

stopCluster(cl)

write_csv(data.frame(res_sim), "qmp_sim_slope_linda.csv")
```

## Visualization

```{r, fig.width=10, fig.height=6.25}
df_ancombc2 = read_csv("../data/qmp/sim_mixed/qmp_sim_slope_ancombc2.csv")
df_linda = read_csv("../data/qmp/sim_mixed/qmp_sim_slope_linda.csv")

simpattern = distinct(df_sim_params, n, diff_prop) %>%
  unite("setting", n:diff_prop, sep = ", ")

df_ancombc2_no_filter = df_ancombc2 %>%
  dplyr::select(X1, X2) %>%
  mutate(method = "ANCOM-BC2 (No Filter)",
         setting = rep(simpattern$setting, each = iter_num))
df_ancombc2_ss_filter = df_ancombc2 %>%
  dplyr::transmute(X1 = X3, X2 = X4) %>%
  mutate(method = "ANCOM-BC2 (SS Filter)",
         setting = rep(simpattern$setting, each = iter_num))
df_linda = df_linda %>%
  mutate(method = "LinDA",
         setting = rep(simpattern$setting, each = iter_num))

df_fig = df_ancombc2_no_filter %>%
  bind_rows(df_ancombc2_ss_filter) %>%
  bind_rows(df_linda) %>%
  separate(setting, c("n", "diff_prop"), ", ")
df_fig$method = factor(df_fig$method, 
                       levels = c("ANCOM-BC2 (No Filter)", 
                                  "ANCOM-BC2 (SS Filter)", 
                                  "LinDA"))

fig_power_slope = df_fig %>%
  ggline(x = "n", y = "X1", add = "mean_se",
         color = "method", xlab = "Sample Size", ylab = "Power", 
         facet.by = "diff_prop", nrow = 1) +
  scale_color_manual(values = c("#E64B35FF", "#4DBBD5FF", "#3C5488FF")) +
  scale_x_discrete(labels = c(10, 20, 30, 50, 100)) +
  guides(color = guide_legend(title = NULL))

fig_fdr_slope = df_fig %>%
  ggline(x = "n", y = "X2", add = "mean_se",
         color = "method", xlab = "Sample Size", ylab = "mdFDR", 
         facet.by = "diff_prop", nrow = 1) +
  scale_color_manual(values = c("#E64B35FF", "#4DBBD5FF", "#3C5488FF")) +
  scale_x_discrete(labels = c(10, 20, 30, 50, 100)) +
  guides(color = guide_legend(title = NULL)) +
  geom_hline(yintercept = 0.05, linetype = "dashed")

fig_slope = ggarrange(fig_fdr_slope, fig_power_slope, 
                      ncol = 1, common.legend = TRUE)

# Simulation summary
df_tab = df_fig %>%
  unite("setting", n:diff_prop, sep = ", ")
df_tab$setting = factor(df_tab$setting, 
                        levels = c(paste0("30, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("60, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("90, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("150, ", c(0.1, 0.2, 0.5, 0.9)),
                                   paste0("300, ", c(0.1, 0.2, 0.5, 0.9))))

tab = df_tab %>% 
  group_by(method, setting) %>% 
  summary_table(summary_template)

addWorksheet(wb, "slope")
writeData(wb, "slope", tab)
```

# Outputs

```{r}
saveWorkbook(wb, file = "../results/outputs/qmp_sim_summ_mixed.xlsx", overwrite = TRUE)

fig = ggarrange(fig_fdr_int, fig_power_int, fig_fdr_slope, fig_power_slope,
                ncol = 1, labels = c("a", "", "b", ""), 
                common.legend = TRUE, legend = "bottom")
ggsave(filename = "../results/figures/supp_qmp_sim_mixed.jpeg", 
       plot = fig, width = 10, height = 10, dpi = 100)
ggsave(filename = "../results/figures/supp_qmp_sim_mixed.pdf", 
       plot = fig, width = 10, height = 10)
```

# Session information

```{r, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```









